Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic

### 2024-03-26 paper.tex
up tex files

\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}

@@

  \usepackage{unicode-math}

@@

@@

%   colorlinks=true,
\urlstyle{same} % disable monospaced font for URLs

@@

\setcounter{secnumdepth}{5}

@@

  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}

\title{Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic}
    \usepackage[]{authblk}
    \setlength{\affilsep}{2em}
    \renewcommand*{\Affilfont}{\normalsize\normalfont}
    \renewcommand\Authsep{, }
    \renewcommand\Authand{ \& }
    \renewcommand\Authands{, \& }
                                        \author[1, 3]{Juan Carlos Castillo}
                                                            \affil[]{Department of Sociology, Universidad de Chile \thanks{\href{mailto:juancastillov@uchile.cl}{\nolinkurl{juancastillov@uchile.cl}}}}
                                                                                \author[2]{Julio Iturra}
                                                            \affil[]{Bremen International Graduate School of Social Sciences, Universität Bremen}
                                                                                \author[3]{Kevin Carrasco}
                                                            \affil[]{Center for Social Conflict and Cohesion Studies}
                                            \date{}

\providecommand{\keywords}[1]
{
  \small
  \textbf{Keywords:} #1
}

@@

\section{Introduction}\label{introduction}

Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated. As the last PISA Report points out, ``Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries'' (\citeproc{ref-oecd_pisa_2023}{OECD, 2023}). Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 (\citeproc{ref-darmodyImpactsCOVID19Control2021}{Darmody et al., 2021}). Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education (\citeproc{ref-leeFairnessPerceptionsEducational2023}{Lee \& Stacey, 2023}). In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.

Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background. In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}; \citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}). This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed (\citeproc{ref-hadjar_meritokratie_2008}{Hadjar, 2008}; \citeproc{ref-mcnamee_meritocracy_2004}{McNamee \& Miller, 2004}; \citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.

The Chilean case is particularly relevant when it comes to the study of inequalities in education. The country is characterized by a high level of economic inequality associated with the privatization and commodification of several social policy domains, among them the educational system (\citeproc{ref-bellei_estudio_2013}{Bellei, 2013}; \citeproc{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et al., 2016}). The neoliberal reforms implemented during and since the military dictatorship (1973-1989) have deepened educational policies based on private subsidies, achievement incentives, selection, and segregation (\citeproc{ref-joikomujicaCuasimercadoEducativoChile2019}{Joiko Mujica, 2019}). Previous research in Chile has emphasized the concept of an ``educational market'' (\citeproc{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et al., 2016}), characterized by selection, competition, and the provision of goods and services associated with payment capacities. Many studies have addressed this problem from economic, political, and sociological perspectives, generating relevant evidence on their academic and social consequences (\citeproc{ref-ramosarellanoEducacionPromesaIncumplida2022}{Ramos Arellano, 2022}). Likewise, massive social mobilizations in the country (particularly in 2006, 2011, and 2018) have had educational inequalities as one of the main targets, putting pressure on governments towards a series of reforms that have advanced free higher education reforms and have diminished the role of private actors in the educational system.

This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework. Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic (\citeproc{ref-breznau_welfare_2021}{Breznau, 2021}; \citeproc{ref-hankivsky_introduction_2022}{Hankivsky et al., 2022}). A pertinent aspect of this discourse is articulated in the works of Mijs (\citeproc{ref-mijs_stratified_2016}{2016}, \citeproc{ref-mijs_visualizing_2018}{2018}), showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises. Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals. Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.

\section{Inequality, beliefs and justification}\label{inequality-beliefs-and-justification}

The justification of economic inequalities is a controversial field of study. As equality is usually considered a synonym of justice in common sense, the idea that someone could be willing to justify inequalities seems (in principle) far from what is considerably reasonable. Nevertheless, and as pointed out by Barry, the whole normative discussion in social justice can be conceived as ``the defensibility of unequal relations between people'' (\citeproc{ref-barry_theories_1989}{1989, p. 3}), whereby empirical research has consistently shown strong variations between individuals and societies regarding inequality justification (\citeproc{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021}; \citeproc{ref-mijs_paradox_2021}{Mijs, 2021}; \citeproc{ref-trump_income_2018}{Trump, 2018}). Both theoretical and empirical approaches have been able to establish a clear difference between inequality and justice, which opens the question of to what extent inequality can be tolerated or even justified.

The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research (\citeproc{ref-Resh2016}{Resh \& Sabbagh, 2016}). From a sociological point of view, the works of Kluegel \& Smith (\citeproc{ref-kluegelBeliefsStratification1981}{1981}) can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice. Theoretically, the authors understand the concept of ``beliefs'' in a ``broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action.'' (\citeproc{ref-kluegelBeliefsStratification1981}{Kluegel \& Smith, 1981, p. 30}). However, the literature has strained the concept of ``beliefs'' by more specific distinctions around the individual and structural dimensions of economic inequality. In this regard, Wegener (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{1987}) has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{Wegener, 1987, p. 2}). This approach established that status positions influence how people perceive things \emph{are} and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust (\citeproc{ref-jassoJusticeEarningsNew1978}{Jasso, 1978}). Additionally, Janmaat (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{2013}) has suggested more detailed conceptual distinctions about how people perceive economic inequality \emph{is}, how they believe or prefer it \emph{should be}, and the process of evaluating such differences as fair.

Theoretically, Liebig \& Sauer (\citeproc{ref-liebigSociologyJustice2016}{2016}) establish three fundamental distinctions in justice attitudes in the field of sociological justice research. First, \emph{order-related attitudes} refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement. Second, \emph{procedural justice attitudes} account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities). Finally, they can be \emph{outcome-related attitudes} which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes). Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles (\citeproc{ref-kluegel_social_1995}{Kluegel et al., 1995}; \citeproc{ref-sonhing_merit_2011}{Son Hing et al., 2011})

An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{Janmaat, 2013}; \citeproc{ref-kluegelBeliefsInequalityAmericans1986}{Kluegel \& Smith, 1986}). The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries (\citeproc{ref-Evans1992}{Evans et al., 1992}; \citeproc{ref-Evans2017}{Evans \& Kelley, 2017}) and greater justification of salary gaps (\citeproc{ref-hadler_why_2005}{Hadler, 2005}; \citeproc{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021}). Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction (\citeproc{ref-jost_psychology_2003}{Jost \& Hunyady, 2003}; \citeproc{ref-vandertoorn_twenty_2014}{van der Toorn \& Jost, 2014}). Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies (\citeproc{ref-trump_when_2020}{Trump, 2020}). As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position (\citeproc{ref-jost_theory_2020}{Jost, 2020}).

Another factor that has been linked to inequality beliefs is the perception of economic inequality (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}; \citeproc{ref-kelleyLegitimationInequalityOccupational1993}{Kelley \& Evans, 1993}; \citeproc{ref-osberg_fair_2006}{Osberg \& Smeeding, 2006}). Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as fair in their generative processes, then they will be considered as just (\citeproc{ref-wegenerIllusionDistributiveJustice1987}{Wegener, 1987}, \citeproc{ref-Wegener1990}{1990}). Thus, both the role of social status and economic inequality perceptions have found support in correlational (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012}; \citeproc{ref-schroder_income_2017}{Schröder, 2017}), as well as in experimental studies that have manipulated inequality perception (\citeproc{ref-iturra_percepcion_2023}{Iturra et al., 2023}; \citeproc{ref-trump_income_2018}{Trump, 2018}). In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality. For instance, studies suggest that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification (\citeproc{ref-Castillo2011}{Castillo, 2011}). Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences (\citeproc{ref-garcia-sanchez_creencias_2022}{García-Sánchez \& De Carvalho Galvão, 2022}). Altogether, these studies have shown that not only socio-structural factors influence what people believe regarding the distribution of economic resources, but their normative stances and perceptions can also determine preferences for economic inequality in diverse domains.

\section{The justification of educational inequality}\label{the-justification-of-educational-inequality}

Studies addressing inequality justification have assessed the support for market ideologies, salary gaps, and distributive preferences. A strand of this research has focused on specific policy areas such as health, pensions, and education, attending to the degree of justification of market-based access to social services. For instance, Lindh (\citeproc{ref-lindhPublicOpinionMarkets2015}{2015}) found that public support for the market distribution of social services is generally low in most countries, implying that most people believe it is unfair for market forces to determine access to basic social services. Similarly, Busemeyer \& Iversen (\citeproc{ref-busemeyer_welfare_2020}{2020}) shows that high-income citizens are less supportive of expanding public spending on universal benefit schemes. Instead, they tend to prefer a public basic insurance scheme, even if it provides relatively more benefits to low-income individuals.

The study of inequality justification in the sphere of education is a rather specific and unattended field. Most studies on educational inequality relate performance indicators (such as grades and standardized test results) to stratification-related variables, whereby few refer specifically to if and how individuals tolerate and even justify educational inequalities. This is probably because in principle it seems counter-intuitive to think that someone would be willing to justify differences in such a sensitive area. This paradox is phrased by Conforti (\citeproc{ref-confortiLegitimationInequalityAmerican1992}{1992}) as ``while equality is embraced as a value, inequality is accepted as a reality'' (p.230).

@@

Studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies. Some perspectives emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about---and more supportive of proposals to close---wealth-based achievement gaps than Black-White or Hispanic-White gaps (\citeproc{ref-valantPoliticsAchievementGaps2016}{Valant \& Newark, 2016}). Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification (\citeproc{ref-stoilovaFairnessEducationalOpportunities2023}{Stoilova \& Ilieva-Trichkova, 2023}). However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.

The perception of economic inequality is one of the factors associated with the justification of educational inequality. As pointed out by Janmaat (\citeproc{ref-janmaatSubjectiveInequalityReview2013}{2013}), justifications are usually confused under the labels of ``beliefs'' and ``attitudes,'' and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized (\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}; \citeproc{ref-sonhing_merit_2011}{Son Hing et al., 2011}). For instance, a recent study by Day \& Norton (\citeproc{ref-dayPerceivedIdealInequality2023}{2023}) on the perception and justification of inequality in university endowments in the U.S. points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality. Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education. Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \& Treisman, 2018})

Besides focusing on the relationship between perception and justification of educational inequality, another group of studies has addressed how the justification of inequality is associated with the support of different distributive principles. In an experiment by Igliozzi et al. (\citeproc{ref-igliozziFairShareEffects2024}{2024}), they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education. They find that a larger system justification (\citeproc{ref-jost_psychology_2003}{Jost \& Hunyady, 2003}) is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance (`Students who place within the highest 50\% of testing scores from the previous year can attend the magnet school'). In the same line, Lee \& Stacey (\citeproc{ref-leeFairnessPerceptionsEducational2023}{2023}) show that performance assessments (in what they call ``neoliberal orientations'') predicted people's fair perceptions of educational inequality.

@@

The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research. The concept of meritocracy, coined by the sociologist Michael Young in his satirical work ``The Rise of the Meritocracy'' (\citeproc{ref-young_rise_1962}{1962}), has a rich history and multifaceted meaning. Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege. However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities (\citeproc{ref-bell_politics_2020}{Bell, 2020}). This would contribute to the justification of inequality in the educational domain (\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}), as individual ability is often inherited, and educational attainment is influenced by family background (\citeproc{ref-goldin_meritocracy_2000}{Goldin, 2000}).

>>>>

\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}

@@

  \usepackage{unicode-math} % this also loads fontspec

@@

\ifPDFTeX\else
  % xetex/luatex font selection

@@

@@

\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \setlength{\parskip}{#2\cslentryspacingunit}

@@

\urlstyle{same}
  colorlinks=true,

\title{Changes in the justification of educational inequalities: The
role of perceptions of inequality and meritocracy during the COVID
pandemic}
\author{}
\date{\vspace{-2.5em}}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Despite the widespread recognition of education as a key factor for
social mobility and equality of opportunities, disparities in this
domain are far from being eliminated. As the last PISA Report points
out, ``Socio-economically disadvantaged students are seven times more
likely than advantaged students to score below Level 2 in mathematics on
average across OECD countries''
(\protect\hyperlink{ref-oecd_pisa_2023}{OECD, 2023}). Furthermore,
empirical evidence from social science research has revealed an
exacerbation of educational inequalities across various domains, such as
family socioeconomic status, gender, ethnicity, and migratory
background, particularly after COVID-19
(\protect\hyperlink{ref-darmodyImpactsCOVID19Control2021}{Darmody et
al., 2021}). Nevertheless, and despite the consistent empirical evidence
on educational inequalities and their consequences, research on
distributive justice indicates remarkable individual differences in the
justification of unequal access to education
(\protect\hyperlink{ref-leeFairnessPerceptionsEducational2023}{Lee \&
Stacey, 2023}). In the present paper, we argue that the perception of
meritocracy in society might play a role in the justification of
economic inequalities in this particular domain.

Meritocracy is defined as a distributive system where individual factors
such as effort and talent are conceived as the main determinants of
individual performance, usually sidelining the role of opportunities and
social background. In recent years, there has been a growing interest in
studying meritocratic beliefs as well as their variations according to
individual and contextual variables
(\protect\hyperlink{ref-mijs_paradox_2021}{Mijs, 2021};
\protect\hyperlink{ref-mijs_belief_2022}{Mijs et al., 2022}). This line
of research has highlighted how meritocratic beliefs are associated with
justifying unequal access to resources and rewards in society, by
legitimizing social hierarchies in which those who already have
resources and advantages are more likely to succeed
(\protect\hyperlink{ref-hadjar_meritokratie_2008}{Hadjar, 2008};
\protect\hyperlink{ref-mcnamee_meritocracy_2004}{McNamee \& Miller,
2004}; \protect\hyperlink{ref-sandel_tyranny_2020}{Sandel, 2020}).
Within this framework, this study focuses on the changes in the
justification of inequalities in access to education in Chile, and to
what extent this phenomenon is associated with the perception of
meritocracy and its variations during the COVID pandemic.

The Chilean case is particularly relevant when it comes to the study of
inequalities in education. The country is characterized by a high level
of economic inequality associated with the privatization and
commodification of several social policy domains, among them the
educational system (\protect\hyperlink{ref-bellei_estudio_2013}{Bellei,
2013};
\protect\hyperlink{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et
al., 2016}). The neoliberal reforms implemented during and since the
military dictatorship (1973-1989) have deepened educational policies
based on private subsidies, achievement incentives, selection, and
segregation (\protect\hyperlink{ref-joiko_cuasimercado_2019}{Joiko,
2019}). Previous research in Chile has emphasized the concept of an
``educational market''
(\protect\hyperlink{ref-corvalanMercadoEscolarLibertad2016}{Corvalán et
al., 2016}), characterized by selection, competition, and the provision
of goods and services associated with payment capacities. Many studies
have addressed this problem from economic, political, and sociological
perspectives, generating relevant evidence on their academic and social
consequences (\protect\hyperlink{ref-ramos_educacion_2022}{Ramos,
2022}). Likewise, massive social mobilizations in the country
(particularly in 2006, 2011, and 2018) have had educational inequalities
as one of the main targets, putting pressure on governments towards a
series of reforms that have advanced free higher education reforms and
have diminished the role of private actors in the educational system.

This research endeavors to dissect the nuanced alterations in the
justifications for educational disparities within the Chilean framework.
Central to our investigation is the premise that the rationale
underpinning these inequalities experiences a significant recalibration
during periods of vulnerability and crises, specifically under the
unprecedented strains of the global health and economic cataclysm
triggered by the COVID-19 pandemic
(\protect\hyperlink{ref-breznau_welfare_2021}{Breznau, 2021};
\protect\hyperlink{ref-hankivsky_introduction_2022}{Hankivsky et al.,
2022}). A pertinent aspect of this discourse is articulated in the works
of Mijs (\protect\hyperlink{ref-mijs_stratified_2016}{2016},
\protect\hyperlink{ref-mijs_visualizing_2018}{2018}), showing the
interplay between meritocratic beliefs and the legitimization of
disparities, as well as the potential erosion of meritocratic principles
amidst crises. Within this framework, we point out that periods of
vulnerability (such as the health and economic crisis resulting from the
COVID-19 pandemic) could weaken the justification of social
inequalities, a change that would be at least in part linked to a
challenge of the meritocratic ideals. Furthermore, we suggest that the
association between meritocratic perception and inequality justification
would be weaker among lower-status groups, as they are the most affected
in critical times.

\hypertarget{inequality-beliefs-and-justification}{%
\section{Inequality, beliefs and
justification}\label{inequality-beliefs-and-justification}}

The justification of economic inequalities is a controversial field of
study. As equality is usually considered a synonym of justice in common
sense, the idea that someone could be willing to justify inequalities
seems (in principle) far from what is considerably reasonable.
Nevertheless, and as pointed out by Barry, the whole normative
discussion in social justice can be conceived as ``the defensibility of
unequal relations between people''
(\protect\hyperlink{ref-barry_theories_1989}{1989, p. 3}), whereby
empirical research has consistently shown strong variations between
individuals and societies regarding inequality justification
(\protect\hyperlink{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021};
\protect\hyperlink{ref-mijs_paradox_2021}{Mijs, 2021};
\protect\hyperlink{ref-trump_income_2018}{Trump, 2018}). Both
theoretical and empirical approaches have been able to establish a clear
difference between inequality and justice, which opens the question of
to what extent inequality can be tolerated or even justified.

The field of studies of subjective economic inequality has been
extensively developed in the social sciences, particularly in the area
of social justice research (\protect\hyperlink{ref-Resh2016}{Resh \&
Sabbagh, 2016}). From a sociological point of view, the works of Kluegel
\& Smith
(\protect\hyperlink{ref-kluegelBeliefsStratification1981}{1981}) can be
considered pioneers in framing a research area comprising perceptions,
beliefs, and justifications in topics such as inequality of
opportunities, poverty, and distributive justice. Theoretically, the
authors understand the concept of ``beliefs'' in a ``broad sense to
refer to the information (veridical or non-veridical) about a phenomenon
that an individual uses as a basis both for inferring other information
and for action.''
(\protect\hyperlink{ref-kluegelBeliefsStratification1981}{Kluegel \&
Smith, 1981, p. 30}). However, the literature has strained the concept
of ``beliefs'' by more specific distinctions around the individual and
structural dimensions of economic inequality. In this regard, Wegener
(\protect\hyperlink{ref-wegenerIllusionDistributiveJustice1987}{1987})
has suggested that perceptions of social status and reward hierarchy are
relative to the observer's status in such hierarchies
(\protect\hyperlink{ref-wegenerIllusionDistributiveJustice1987}{Wegener,
1987, p. 2}). This approach established that status positions influence
how people perceive things \emph{are} and, in turn, these perceptions
anchor how people evaluate this inequality as just or unjust
(\protect\hyperlink{ref-jassoJusticeEarningsNew1978}{Jasso, 1978}).
Additionally, Janmaat
(\protect\hyperlink{ref-janmaatSubjectiveInequalityReview2013}{2013})
has suggested more detailed conceptual distinctions about how people
perceive economic inequality \emph{is}, how they believe or prefer it
\emph{should be}, and the process of evaluating such differences as
fair.

Theoretically, Liebig \& Sauer
(\protect\hyperlink{ref-liebigSociologyJustice2016}{2016}) establish
three fundamental distinctions in justice attitudes in the field of
sociological justice research. First, \emph{order-related attitudes}
refer to individual motives that explain preferences for resource
allocation based on the principles of equality, need, or entitlement.
Second, \emph{procedural justice attitudes} account for the degree of
acceptance of unfavorable outcomes under conditions in which the
procedure under which those decisions were made is considered fair
(e.g., opportunities). Finally, they can be \emph{outcome-related
attitudes} which refer primarily to the economic resources that a person
currently obtains and the set of collective commitments associated with
possessing those resources (e.g., transfers or taxes). Within this
framework, we conceptualize the justification of inequality in the first
area of order-related attitudes, as it expresses preferences for
resource allocation that can be associated with concepts that reflect
different distributive principles
(\protect\hyperlink{ref-kluegel_social_1995}{Kluegel et al., 1995};
\protect\hyperlink{ref-sonhing_merit_2011}{Son Hing et al., 2011})

An important finding in the area of order-related attitudes is that
beliefs about inequality are stratified by social status
(\protect\hyperlink{ref-janmaatSubjectiveInequalityReview2013}{Janmaat,
2013};
\protect\hyperlink{ref-kluegelBeliefsInequalityAmericans1986}{Kluegel \&
Smith, 1986}). The evidence suggests that advantaged positions in terms
of socioeconomic status, such as income, education, or occupational
prestige, are associated with preferences for less egalitarian class
imageries (\protect\hyperlink{ref-Evans1992}{Evans et al., 1992};
\protect\hyperlink{ref-Evans2017}{Evans \& Kelley, 2017}) and greater
justification of salary gaps
(\protect\hyperlink{ref-hadler_why_2005}{Hadler, 2005};
\protect\hyperlink{ref-kelley_legitimate_2021}{Kelley \& Evans, 2021}).
Nevertheless, and in contrast to this perspective, it has been argued
that human beings are intrinsically motivated to justify the system they
inhabit, even when the information available to them points in the
opposite direction (\protect\hyperlink{ref-jost_psychology_2003}{Jost \&
Hunyady, 2003}; \protect\hyperlink{ref-vandertoorn_twenty_2014}{van der
Toorn \& Jost, 2014}). Individuals would be generally disposed to
justify the economic status quo regardless of self- and group economic
interest because their support partially depends on their beliefs in
system-justifying ideologies
(\protect\hyperlink{ref-trump_when_2020}{Trump, 2020}). As a result,
what has been suggested is that low-status might perceive the status quo
as just as they are more strongly attached to dominant ideologies, while
high-status individuals are more prone to justify the system to a
greater extent because it favors their interests and dominant position
(\protect\hyperlink{ref-jost_theory_2020}{Jost, 2020}).

Another factor that has been linked to inequality beliefs is the
perception of economic inequality
(\protect\hyperlink{ref-castillo_perception_2022}{Castillo et al.,
2022};
\protect\hyperlink{ref-kelleyLegitimationInequalityOccupational1993}{Kelley
\& Evans, 1993}; \protect\hyperlink{ref-osberg_fair_2006}{Osberg \&
Smeeding, 2006}). Studies in this matter have argued that perceived
economic inequality functions as an anchor for the justifications of
inequality because if these inequalities are indeed seen as fair in
their generative processes, then they will be considered as just
(\protect\hyperlink{ref-wegenerIllusionDistributiveJustice1987}{Wegener,
1987}, \protect\hyperlink{ref-Wegener1990}{1990}). Thus, both the role
of social status and economic inequality perceptions have found support
in correlational
(\protect\hyperlink{ref-Castillo2012a_justice}{Castillo, 2012};
\protect\hyperlink{ref-schroder_income_2017}{Schröder, 2017}), as well
as in experimental studies that have manipulated inequality perception
(\protect\hyperlink{ref-iturra_percepcion_2023}{Iturra et al., 2023};
\protect\hyperlink{ref-trump_income_2018}{Trump, 2018}). In addition,
normative or value-driven mechanisms can also explain to what extent
people justify inequality. For instance, studies suggest that justice
ideologies such as individualism or egalitarianism can mediate the
influence of individual status on income inequality justification
(\protect\hyperlink{ref-Castillo2011}{Castillo, 2011}). Likewise, system
justification ideologies can strengthen the influence of socioeconomic
status on economic inequality preferences
(\protect\hyperlink{ref-garcia-sanchez_creencias_2022}{García-Sánchez \&
De Carvalho, 2022}). Altogether, these studies have shown that not only
socio-structural factors influence what people believe regarding the
distribution of economic resources, but their normative stances and
perceptions can also determine preferences for economic inequality in
diverse domains.

\hypertarget{the-justification-of-educational-inequality}{%
\section{The justification of educational
inequality}\label{the-justification-of-educational-inequality}}

Studies addressing inequality justification have assessed the support
for market ideologies, salary gaps, and distributive preferences. A
strand of this research has focused on specific policy areas such as
health, pensions, and education, attending to the degree of
justification of market-based access to social services. For instance,
Lindh (\protect\hyperlink{ref-lindhPublicOpinionMarkets2015}{2015})
found that public support for the market distribution of social services
is generally low in most countries, implying that most people believe it
is unfair for market forces to determine access to basic social
services. Similarly, Busemeyer \& Iversen
(\protect\hyperlink{ref-busemeyer_welfare_2020}{2020}) shows that
high-income citizens are less supportive of expanding public spending on
universal benefit schemes. Instead, they tend to prefer a public basic
insurance scheme, even if it provides relatively more benefits to
low-income individuals.

The study of inequality justification in the sphere of education is a
rather specific and unattended field. Most studies on educational
inequality relate performance indicators (such as grades and
standardized test results) to stratification-related variables, whereby
few refer specifically to if and how individuals tolerate and even
justify educational inequalities. This is probably because in principle
it seems counter-intuitive to think that someone would be willing to
justify differences in such a sensitive area. This paradox is phrased by
Conforti
(\protect\hyperlink{ref-confortiLegitimationInequalityAmerican1992}{1992})
as ``while equality is embraced as a value, inequality is accepted as a
reality'' (p.230).

@@

Studies have addressed the justification of educational inequality from
different conceptual perspectives and methodological strategies. Some
perspectives emphasize the justification of differences between specific
groups, finding for instance that Americans are more concerned
about---and more supportive of proposals to close---wealth-based
achievement gaps than Black-White or Hispanic-White gaps
(\protect\hyperlink{ref-valantPoliticsAchievementGaps2016}{Valant \&
Newark, 2016}). Other studies focus on the association between status
differences and inequality justification in education, based on the
rational premise that those who are worst-off would tend to show lower
levels of justification
(\protect\hyperlink{ref-stoilovaFairnessEducationalOpportunities2023}{Stoilova
\& Ilieva-Trichkova, 2023}). However, most of the research in this area
goes beyond status differences, focusing on how subjective factors such
as perceptions and beliefs are related to different levels of
justification.

The perception of economic inequality is one of the factors associated
with the justification of educational inequality. As pointed out by
Janmaat
(\protect\hyperlink{ref-janmaatSubjectiveInequalityReview2013}{2013}),
justifications are usually confused under the labels of ``beliefs'' and
``attitudes,'' and their distinction from perceptions is relevant as it
helps state that justifying a certain level of inequality requires
taking into account which level of (perceived) inequality is first and
foremost being recognized
(\protect\hyperlink{ref-castillo_multidimensional_2023}{Castillo et al.,
2023}; \protect\hyperlink{ref-sonhing_merit_2011}{Son Hing et al.,
2011}). For instance, a recent study by Day \& Norton
(\protect\hyperlink{ref-dayPerceivedIdealInequality2023}{2023}) on the
perception and justification of inequality in university endowments in
the U.S. points out that individuals tend to underestimate the magnitude
of existing inequality while simultaneously desiring greater equality.
Furthermore, through an experiment, they demonstrate that information
about inequality in university endowments (i.e., manipulating inequality
perception) increases the perception of injustice in education. Results
of this study are in line with the consistent evidence on biases related
to the underestimation of economic inequality
(\protect\hyperlink{ref-castillo_perception_2022}{Castillo et al.,
2022}; \protect\hyperlink{ref-gimpelson_misperceiving_2018}{Gimpelson \&
Treisman, 2018})

Besides focusing on the relationship between perception and
justification of educational inequality, another group of studies has
addressed how the justification of inequality is associated with the
support of different distributive principles. In an experiment by
Igliozzi et al.
(\protect\hyperlink{ref-igliozziFairShareEffects2024}{2024}), they
establish how the support for different justice principles (equity,
equality, and need) is related to support for redistributive policies in
domains such as education. They find that a larger system justification
(\protect\hyperlink{ref-jost_psychology_2003}{Jost \& Hunyady, 2003}) is
related to higher support for the principles of equity and equality
(over need) in distributing educational outcomes, where equity was
measured in terms of access to education related to performance
(`Students who place within the highest 50\% of testing scores from the
previous year can attend the magnet school'). In the same line, Lee \&
Stacey
(\protect\hyperlink{ref-leeFairnessPerceptionsEducational2023}{2023})
show that performance assessments (in what they call ``neoliberal
orientations'') predicted people's fair perceptions of educational
inequality.

@@

The relationship between performance and outputs brings us to the
concepts of merit and meritocracy, key in educational justice research.
The concept of meritocracy, coined by the sociologist Michael Young in
his satirical work ``The Rise of the Meritocracy''
(\protect\hyperlink{ref-young_rise_1962}{1962}), has a rich history and
multifaceted meaning. Young's term was initially used to critique a
hypothetical future society where social and economic positions were
entirely determined by individuals' merits, talents, and efforts,
ostensibly eliminating the influence of inherited privilege. However,
over time, the term has evolved to describe a social system where
advancement and success are perceived as directly linked to one's
abilities and achievements, sidelining inequalities of opportunities
(\protect\hyperlink{ref-bell_politics_2020}{Bell, 2020}). This would
contribute to the justification of inequality in the educational domain
(\protect\hyperlink{ref-sandel_tyranny_2020}{Sandel, 2020}), as
individual ability is often inherited, and educational attainment is
influenced by family background
(\protect\hyperlink{ref-goldin_meritocracy_2000}{Goldin, 2000}).
